pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-05-13 bmc_article.tex

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score across subpopulations. A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datasets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). PCA-based selection did not appear superior to feature-wise filters. We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering tended improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g. FACS-sorted datasets). Indeed, the datasets where doublet removal did not have a clear positive impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact among the Smart-seq datasets (Koh and Kumar). When performed with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datasets, possibly owing to its regularization.  In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 20-21). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 20-22). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 20-21). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also tended to lead to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms. We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.

We compared the various estimates of in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). These truth-based estimates where then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

\subsection*{Overview of interactions between steps}

We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions. Analysis of variance (after correcting for baseline differences between datasets and the resolution parameter) attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality, followed by normalization, and very little to filtering (Supplementary Figure 28). Because the choice of dimensions impacts the number of clusters, we repeated the analysis of variance among the combinations of methods (up to clustering) that yielded the true number of clusters. We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.

We next investigated whether the benefits of some methods where conditional on the choices at other steps. Although analysis of variance identified many weak but significant interactions, in general we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30). For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29). Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).

@@

\subsection*{Application of \textif{pipeComp} to a different context}

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with 1-3 variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

\subsection*{Practical recommendations}

On the basis of our findings, we can make a number of practical analysis recommendations, also summarized in Supplementary Figure 34:

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions - see for instance our 'default' filters in the methods) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics. Applying this framework to scRNAseq clustering enabled us to make clear recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34). We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance. We further showed that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.

@@

\subsection*{Overall run for interactions across steps}

>>>>

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one {\color{red}(Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail).} Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

{\color{red}We next investigated the impact of several filtering methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{veryStringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel})}. An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored the mean F1 score {\color{red}across subpopulations. A first observation was that, although more stringent filtering tended to be associated with an increase in accuracy, it tended to plateau and could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. In contrast, too lenient filtering (or lack thereof) can lead to misclassified cells, as was the case for the Zhengmix4 datasets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). PCA-based selection did not appear superior to feature-wise filters. We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering tended to improve accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (e.g. FACS-sorted datasets). Indeed, the datasets where doublet removal did not have a clear positive impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.}

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI, Figure \ref{fig:norm}B-C). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations{\color{red}, and that some datasets (e.g. Zhengmix8eq and simMix1) and subpopulations were consistently more challenging across all methods. An exception was \textit{scVI}, which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy (Supplementary Figure 13; \textit{scVI} was run on a reduced set of pipelines). Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact among the Smart-seq datasets (Koh and Kumar). When combined with \textit{SCTransform}, regressing out covariates using tended to have neither a positive or negative effect in droplet-based datasets, possibly owing to its regularization.  In general, \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).}

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated ({\color{red}Pearson correlation >0.9,} Supplementary Figure 19A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 20-21). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 20-22). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 20-21). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also {\color{red}tended to lead} to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). {\color{red}Overall, weighting the components by their explained variance tended to increase silhouette width (Supplementary Figure 23A) and, in most datasets, clustering accuracy (Supplementary} Figure 23B). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. {\color{red}Its performance was however the worst both in terms of silhouette width and clustering accuracy (Supplementary Figure 15), and it only performed well in some of the least challenging datasets (mixology10x5cl and simMix2).}

@@

Different approaches have been proposed to select the appropriate number of dimensions, from the visual identification of an {\color{red}inflexion point in the curve of the variance explained by each component (`elbow' method), to more complex algorithms.} We evaluated the performance of dimensionality estimators implemented in the \textit{intrinsicDimension} package \cite{johnssonDimensionality2015}, as well as common procedures such as the `elbow` method, some scRNAseq-specific methods such as the JackStraw procedure \cite{ChungJackstraw2015} or \textit{scran}'s \texttt{denoisePCA} \cite{LunScran2016}, and the recent application of Fisher Separability analysis \cite{AlberganteSepar2019}.

{\color{red}We compared the various estimates in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component (based on Seurat's weighted PCA space) that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). These truth-based estimates were then compared to those of the above dimensionality estimation methods (Figure \ref{fig:dim}B).} Of note, the methods differ widely in compute time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model {\color{red}(\textit{maxLikGlobal}, using 10 or 20 nearest neighbors, although estimates were relatively stable across various values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated challenging subpopulations (Figure \ref{fig:dim}C) and tended to result in the best clustering accuracy (Figure \ref{fig:dim}D). This method systematically selected many more components than were associated with the subpopulations in Figure \ref{fig:dim}A, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An important implication is that the commonly-used elbow method most likely underestimates dimensionality.}

@@

{\color{red}\subsection*{Overview of interactions between steps}

We selected the most important alternatives for each parameter (see Methods) in order to explore their interactions. After accounting, in the model, for baseline differences between datasets and the impact of the resolution parameter, ANOVA attributed the largest proportion of variance in the main cluster metrics to the choice of dimensionality reduction, followed by normalization, and very little to filtering (Supplementary Figure 28A). Because the choice of dimension reduction impacts the number of clusters, we repeated the analysis of variance in ARI among the combinations of parameters that yielded the true number of clusters (Supplementary Figure 28B). We observed that although normalization was the most important parameter among these analyses, the choice of dimensionality came closely behind (Supplementary Figure 28B), suggesting that more attention should be paid to this relatively neglected step.

We next investigated whether the benefits of some methods were conditional on the choices at other steps. Although analysis of variance identified many weak but significant interactions, we found that the benefits brought by performant methods were robust to differences in other parts of the pipeline (Supplementary Figures 29-30). For instance, we confirmed that doublet removal and \textit{SCtransform} normalization were beneficial independently of other steps, and that, again independendently of other steps, more stringent filtering did not yield any major improvement over lenient filtering (Supplementary Figure 29). Similarly, deviance-based feature selection outcompeted \textit{Seurat}'s default selection method independently of changes at other steps (Supplementary Figure 30).}

@@

{\color{red}\subsection*{Application of \textif{pipeComp} to a different context}

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 32A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 32A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 32-33). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with 1-3 variables. These results therefore suggest that these methods are robust and can be widely applied.}

@@

\subsection*{{\color{red}Practical} recommendations}

On the basis of our findings, we can make a number of \textcolor{red}{practical} analysis recommendations, also summarized in Supplementary Figure 34:

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions \textcolor{ref}{- see for instance our 'default' filters in the methods}) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{SCTransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances\textcolor{red}{, although none of these methods was beneficial over all subpopulations}.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the \textcolor{red}{10 or }20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

\textit{pipeComp} is a flexible R framework for evaluating methods and parameter combinations in a complex pipeline, computing on-the-fly multilevel evaluation metrics. Applying this framework to scRNAseq clustering enabled us to make {\color{red}clear} recommendations on the steps of filtering, normalization, feature selection, denoising, dimensionality reduction and clustering (Supplementary Figure 34). We demonstrate how a diversity of multilevel metrics can be more robust, more sensitive, and more nuanced than simply evaluating the final clustering performance. We {\color{red}further showed} that the \textit{pipeComp} framework can be applied to extend the current field of benchmarking, as well as to investigate pipelines other domains with multi-step computational analyses.

@@

{\color{red}\subsection*{Overall run for interactions across steps}
